Using physical simulations to predict children's and adults' preference for physical interactions

### 2021-02-01 analysis/model_analysis_v2.Rmd
added model plot

load_data <- function(fname) {
  raw <- read.table(here(paste0("models/",fname,".csv")), sep=',')
  cnames <- raw[1,]
  mdat <- raw[2:nrow(raw),]
  names(mdat) = cnames
  for(c in 1:(ncol(mdat)-3)) {
    mdat[,c] = as.numeric(mdat[,c])
    if(min(mdat[,c]) < 0) mdat[,c] = mdat[,c] + min(mdat[,c])
    #print(paste("shifting",c,"so minimum is 0"))
  }
  mdat$trial = rep(1:20, each=2)
  # right now, each trial is 2 rows, 
  # uniquely identified with drop_object x condition
  #mdat %>% distinct(drop_object, condition)

@@

# just 100 drops per trial
mdat100 <- load_data("model_stats")

@@

@@

get_model_target_choice_props <- function(feat_df, hum_df, beta) {
  hum_df <- hum_df %>% arrange(relation, drop) # same order as model df
  hum_df$Model_targ_prop = 0
  for(i in 1:nrow(hum_df)) {
    mod_cols = which(!is.na(feat_df[i,1:ncol(feat_df)]))[3:4] # CHECK THIS
    mod_choice = softmax(feat_df[i,mod_cols], beta=beta)
    hum_df[i,]$Model_targ_prop = as.numeric(mod_choice[hum_df[i,]$target] / sum(mod_choice))
  }
  return(hum_df)
}

@@

Could also fit separately to contain and support trials with MSE - r objective...

@@

```{r}
ch_feat <- all_trials_mse %>% filter(Age=="children") %>%
  arrange(mse)

@@

@@

ad <- plot_model_preds("support_probability", all_trials_mse, adult_trial_agg, age="adults", mdat)
ch <- plot_model_preds("support_probability", all_trials_mse, child_trial_agg, age="children", mdat)

@@

>>>>

load_data <- function(fname) {
  raw <- read.table(here(paste0("models/",fname,".csv")), sep=',')
  cnames <- raw[1,]
  mdat <- raw[2:nrow(raw),]
  names(mdat) = cnames
  for(c in 1:(ncol(mdat)-3)) {
    mdat[,c] = as.numeric(mdat[,c])
    if(min(mdat[,c]) < 0) {
      mdat[,c] = mdat[,c] + min(mdat[,c])
      print(paste("shifting",c,"so minimum is 0"))
    }
  }
  mdat$trial = rep(1:20, each=2)
  # right now, each trial is 2 rows, 
  # uniquely identified with drop_object x condition
  #mdat %>% distinct(drop_object, condition)

@@

# just 100 drops per trial
#mdat100 <- load_data("model_stats")

@@

# model split-half correlations and stds
mdat_cor <- read_csv(here("models/model_stats250_corr.csv")) 
# avg_len NA, but all others ~.997 - .999
#View(mdat_cor)

mdat_std <- load_data("model_stats250_std")
mdat_std$drop = mdat$drop
mdat_std$target = mdat$target
mdat_std$relation = mdat$relation
mdat_std$trial = mdat$trial

@@

get_model_target_choice_props <- function(feat_df, hum_df, beta) {
  hum_df <- hum_df %>% arrange(relation, drop) # same order as model df
  hum_df$Model_targ_prop = 0
  for(i in 1:nrow(hum_df)) {
    mod_cols = which(!is.na(feat_df[i,1:ncol(feat_df)]))[3:4] 
    mod_choice = softmax(feat_df[i,mod_cols], beta=beta)
    hum_df[i,]$Model_targ_prop = as.numeric(mod_choice[hum_df[i,]$target] / sum(mod_choice))
  }
  return(hum_df)
}

@@

@@

```{r}
get_model_error_bars <- function(feat_name, mdat_std) {
  # have std per drop, so we avg per trial (?) 
  # (could get SEM for N=250 drops ?)
  mdat_std <- mdat_std %>% arrange(relation, drop)
  sd_trial <- rowMeans(matrix(mdat_std[,feat_name], nrow=20, byrow=T))
  return(sd_trial)
}

ch_feat <- all_trials_mse %>% filter(Age=="children") %>%
  arrange(mse)

@@

ad_df = get_model_target_choice_props(get_feature_df(mdat, "support_probability"), adult_trial_agg, ad_feat[2,]$beta)
ad_df$model_sd = get_model_error_bars("support_probability", mdat_std)

ch_df = get_model_target_choice_props(get_feature_df(mdat, "support_probability"), child_trial_agg, ch_feat[1,]$beta)
ch_df$model_sd = get_model_error_bars("support_probability", mdat_std)

save(ad_df, ch_df, file=here("models/support_probability_fit_df.RData"))

@@

ad <- plot_model_preds("support_probability", all_trials_mse, adult_trial_agg, age="adults", mdat)
ch <- plot_model_preds("support_probability", all_trials_mse, child_trial_agg, age="children", mdat)
# ToDo: add error bars to each trial for model (from mdat_std) and from humans...

@@

### How much of the total explainable variance (e.g. relative to measurement noise) is explained by the best model for the adult data?

```{r}
md <- get_model_target_choice_props(get_feature_df(mdat, feat_name), hum_data, this_dat$beta)

### 2021-02-01 paper/paper2021.Rmd
added model plot

```{r adults-vs-children, fig.env = "figure", fig.pos = "h", fig.align="center", fig.width=3.5, fig.height=3.75, set.cap.width=T, num.cols.cap=1, fig.cap = "Comparison of children's vs. adults' preferences on each trial (dot), with bootstrapped 95\\% confidence intervals."}
# add vert and horiz error bars 
combo_dat %>% group_by(drop, target, relation, ID) %>%
  group_by(drop, target, relation, AgeGroup) %>%
  tidyboot_mean(chose_target, na.rm=T) %>%
  select(-n, -empirical_stat) %>%
  pivot_wider(names_from = AgeGroup, values_from = c(mean, ci_lower, ci_upper)) %>%
  ggplot(aes(x=mean_adult, y=mean_child, group=relation, color=relation)) + 
    geom_point(alpha=.6) + 
    geom_errorbar(aes(ymin=ci_lower_child, ymax=ci_upper_child), alpha=.6) +
    geom_errorbarh(aes(xmin=ci_lower_adult, xmax=ci_upper_adult), alpha=.6) + 
    theme_bw() + ylim(0,1) + xlim(0,1) + 
    xlab("Adults' Choice Proportions") + ylab("Children's Choice Proportions") + 
    geom_abline(xintercept=0, slope=1, linetype="dashed") +
    geom_smooth(method = "lm") +
    theme(legend.position = "bottom")
  #geom_text(aes(label=target)) #

@@

```{r load-model-data}
load(here("models/model_data.RData")) # mdat, mdat_eng, fits
```
Using a physics engine and 3D models of the objects that were shown to participants, we constructed a simulation of each trial of the experiment and attempted to predict children's and adults' choices using a variety of heuristic models of the simulated physical interactions.
For each trial's two possible drop object choices, we ran 250 simulated drops, and for each drop calculated a variety of features measuring the state of the model after the drop was completed.
We selected features that we thought may provide good metrics for what people would found interesting. 
For example, we collected for each drop the mean length of time before both objects come to rest (*avg_len*), as well as the standard deviation (SD) of that time (*len_std*).
Some features were calculated separately for drop and target objects, such as: the SD of the object's final position (*obj_final_position_std_objects*) as well as the inverse of that (*obj_final_position_invstd_objects*), the mean final distance from the drop location (*avg_final_radius_objects*), the maximum final distance from the drop location (*avg_max_radius_objects*), the ... (*max_radius_std_objects*), ... (*normed_velocity_std_after_first_collision_objects*)
<!-- support_probability, support_std, support_response_linearity_r, support_response_linearity_pv  -->
<!--len_inverse_sharpe_ratio, support_response_sharpness_C=1 -->
[ToDo: describe *skillsiness*]
We measure each of these features in 250 simulations of each possible drop in the experiment, and then use the resulting metrics to generate relative preferences for each of the drop choices on each trial. 
Finally, we compare how well each of these simple heuristic models account for children's and adults' drop preferences on each trial.

```{r}
# plot the best feature for children, adults, 
```

@@

<!-- moved from old Intro: rewrite to discuss Haber et al. more -->
Thexe results will serve as a test set for evaluating computational theories of curiosity, which in turn are inspired by classic descriptive theories positing that curiosity-driven behavior was the consequence of attentional capture by novel stimuli [e.g., @Berlyne1954; @Fantz1964], as well as theories providing qualitative explanations for curiosity-driven behavior as the result of learners' preferences for stimuli that are "moderately discrepant" in relation to their current knowledge state, and thereby provide opportunities to learn [@Kinney1976]. 
Recent studies have provided empirical support for this basic idea: 7- to 8-month-old infants are sensitive to stimulus complexity and prefer to look at moderately complex stimuli rather than simple or highly-complex stimuli [@Kidd2012; @Kidd2014].
According to these theories, curiosity-driven behavior reflects a gap between the learner's knowledge and the state of the world [@Loewenstein1994], and thus they predict that as a learner gains additional knowledge, their preferences will shift toward more complex stimuli [@Dember1957].
Although these theories have provided important qualitative insights, one of their major limitations is that they do not provide precise ways to characterize a learner's current knowledge state, how physical states of the world are represented in their mind, nor how discrepancies between a learner's knowledge and such states are compared.  
Our approach leverages theory and insights from reinforcement learning (RL), including the value of mechanisms that equip RL agents with "intrinsic motivation" — a drive to explore the state space even when rewards are sparse or absent, and favor actions with uncertain outcomes, because they may result in the discovery of new policies with high expected values [@Schmidhuber2010].
Our modeling approach is resonant with recent work that has instantiated such intrinsic-motivation mechanisms in robots to help them learn robust ways to predict physical events in the world world [@Oudeyer2007; @Oudeyer2013], although this work typically involves pretraining models on a separate physical prediction task before implementing curiosity-driven learning, and has not yet been directly compared to human behavior in the same prediction tasks.

>>>>

```{r adults-vs-children, fig.env = "figure", fig.pos = "h", fig.align="center", fig.width=3.5, fig.height=3.75, set.cap.width=T, num.cols.cap=1, fig.cap = "Comparison of children's vs. adults' preferences on each trial (dot), with bootstrapped 95\\% confidence intervals."}
# add vert and horiz error bars 
combo_dat %>% group_by(drop, target, relation, ID) %>%
  group_by(drop, target, relation, AgeGroup) %>%
  tidyboot_mean(chose_target, na.rm=T) %>%
  select(-n, -empirical_stat) %>%
  pivot_wider(names_from = AgeGroup, values_from = c(mean, ci_lower, ci_upper)) %>%
  ggplot(aes(x=mean_adult, y=mean_child, group=relation, color=relation)) + 
    geom_point(alpha=.6) + 
    geom_errorbar(aes(ymin=ci_lower_child, ymax=ci_upper_child), alpha=.6) +
    geom_errorbarh(aes(xmin=ci_lower_adult, xmax=ci_upper_adult), alpha=.6) + 
    theme_bw() + ylim(0,1) + xlim(0,1) + 
    xlab("Adults' Choice Proportions") + ylab("Children's Choice Proportions") + 
    geom_abline(intercept=0, slope=1, linetype="dashed") +
    geom_smooth(method = "lm") +
    theme(legend.position = "bottom")
  #geom_text(aes(label=target)) #

@@

```{r load-model-data}
load(here("models/model_data.RData")) # mdat, all_trials_mse, all_trials_mse_r, etc.
load(here("models/support_probability_fit_df.RData")) # ad_df, ch_df - model (w SD) vs. human (need 95%CIs)

mdat <- ad_df %>% mutate(AgeGroup="adult") %>% 
  bind_rows(ch_df %>% mutate(AgeGroup="child") %>% select(-n)) %>% 
  mutate(mod_ci_lower = Model_targ_prop - 1.96*model_sd,
         mod_ci_upper = Model_targ_prop + 1.96*model_sd) # 95% should fall within +/-1.96 SDs of mean
# now just need 95% CIs for human data
```
Using a physics engine and 3D models of the objects that were shown to participants, we constructed a simulation of each trial of the experiment and attempted to predict children's and adults' choices using a variety of heuristic models of the simulated physical interactions.
For each trial's two possible drop object choices, we ran 250 simulated drops, and for each drop calculated a variety of features measuring the state of the model after the drop was completed.
We selected features that we thought may provide good metrics for what people would found interesting. 
For example, we collected for each drop the mean length of time before both objects come to rest (*avg_len*), as well as the standard deviation (SD) of that time (*len_std*).
Some features were calculated separately for drop and target objects, such as: the SD of the object's final position (*obj_final_position_std_objects*) as well as the inverse of that (*obj_final_position_invstd_objects*), the mean final distance from the drop location (*avg_final_radius_objects*), the maximum final distance from the drop location (*avg_max_radius_objects*), the ... (*max_radius_std_objects*), ... (*normed_velocity_std_after_first_collision_objects*)
<!-- support_probability, support_std, support_response_linearity_r, support_response_linearity_pv  -->
<!--len_inverse_sharpe_ratio, support_response_sharpness_C=1 -->
[ToDo: describe *skillsiness*]
We measure each of these features in 250 simulations of each possible drop in the experiment, and then use the resulting metrics to generate relative preferences for each of the drop choices on each trial. 
Finally, we compare how well each of these simple heuristic models account for children's and adults' drop preferences on each trial.

```{r human-vs-model-relational-choices, fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.25, fig.height=5.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Comparison of model's vs. participants' choices for each type of trial, with bootstrapped 95\\% confidence intervals."}
pos = position_dodge(width = .9)

hum_ag <- combo_dat %>% group_by(drop, target, relation, ID) %>%
  group_by(drop, target, relation, AgeGroup) %>%
  tidyboot_mean(chose_target, na.rm=T) %>%
  select(-n, -empirical_stat)

hum_mod <- hum_ag %>% left_join(mdat)

hum_mod %>%
  ggplot(aes(x=mean, y=Model_targ_prop, group=relation, color=relation)) + 
    geom_point(alpha=.6) + 
    geom_errorbar(aes(ymin=mod_ci_lower, ymax=mod_ci_upper), alpha=.6) +
    geom_errorbarh(aes(xmin=ci_lower, xmax=ci_upper), alpha=.6) + 
    facet_wrap(~ AgeGroup, nrow=2) +
    theme_bw() + ylim(0,1) + xlim(0,1) + 
    xlab("Human Choice Proportions") + ylab("Model Choice Proportions") + 
    geom_abline(intercept=0, slope=1, linetype="dashed") +
    geom_smooth(method = "lm") +
    theme(legend.position = "top")

@@

<!-- moved from old Intro: rewrite to discuss Haber et al. more -->
These results will serve as a test set for evaluating computational theories of curiosity, which in turn are inspired by classic descriptive theories positing that curiosity-driven behavior was the consequence of attentional capture by novel stimuli [e.g., @Berlyne1954; @Fantz1964], as well as theories providing qualitative explanations for curiosity-driven behavior as the result of learners' preferences for stimuli that are "moderately discrepant" in relation to their current knowledge state, and thereby provide opportunities to learn [@Kinney1976]. 
Recent studies have provided empirical support for this basic idea: 7- to 8-month-old infants are sensitive to stimulus complexity and prefer to look at moderately complex stimuli rather than simple or highly-complex stimuli [@Kidd2012; @Kidd2014].
According to these theories, curiosity-driven behavior reflects a gap between the learner's knowledge and the state of the world [@Loewenstein1994], and thus they predict that as a learner gains additional knowledge, their preferences will shift toward more complex stimuli [@Dember1957].
Although these theories have provided important qualitative insights, one of their major limitations is that they do not provide precise ways to characterize a learner's current knowledge state, how physical states of the world are represented in their mind, nor how discrepancies between a learner's knowledge and such states are compared.  
Our approach leverages theory and insights from reinforcement learning (RL), including the value of mechanisms that equip RL agents with "intrinsic motivation" — a drive to explore the state space even when rewards are sparse or absent, and favor actions with uncertain outcomes, because they may result in the discovery of new policies with high expected values [@Schmidhuber2010].
Our modeling approach is resonant with recent work that has instantiated such intrinsic-motivation mechanisms in robots to help them learn robust ways to predict physical events in the world world [@Oudeyer2007; @Oudeyer2013], although this work typically involves pretraining models on a separate physical prediction task before implementing curiosity-driven learning, and has not yet been directly compared to human behavior in the same prediction tasks.
